A general and efficient representation of Ancestral Recombination Graphs

### 2022-04-07 paper.tex
Make node labels lowercase
And add subfigure labels

% NOTE: it might be helpful to have a drawing of the ARG process
% in-progress here to help us talk about it. What is the full state
% that is remembered as we go through the process? It does help to think
% about the process as choosing edges to merge, not nodes.
% We could also refer to Fig.~\ref{fig-arg-data-structure} if we want a simple picture
% to go with this
In the Griffiths formulation, each edge in the graph corresponds to an extant
lineage and nodes are events in the process (the initial $n$ leaf nodes are
``sampling'' events). Common ancestor events occur at rate $\binom{k}{2}$ when there
are $k$ lineages present. We choose two lineages (edges) uniformly, and merge them
into a common ancestor lineage. Recombination events happen at a rate of
$k \rho$. Here, we choose a lineage (edge) uniformly, and a
breakpoint $0 < x < m$ uniformly on its genome. We terminate the edge at a
node, record the breakpoint and start two new edges from this node. The process
then continues until there is only one lineage left (the Grand Most Recent
Common Ancestor, GMRCA), which is guaranteed to
happen in finite time because of the linear vs quadratic rates of branching
and coalescing. Later, we will show how the graph structure and breakpoints
associated with
recombination nodes provides sufficient information to later recover the marginal
trees and calculate sampling probabilities. However, this process traces out
a larger graph than the Hudson algorithm (see the ``\nameref{Ancestry_resolution}'' section),
and is consequently considerably less efficient \citep{wiuf1999ancestry}.

@@

\section*{The Event ARG data structure}
% When did we move from the ARG is a process to the ARG is a data structure?
% What was the historical progression from one to the other?

@@

% What do we mean by an EARG, informally?
The Griffiths ARG encoding is an elegant and mathematically
economical description of the coalescent with recombination, whose
fundamental units are the  \emph{events} that correspond directly
to those that occur in the associated stochastic process. We will refer
to this ARG encoding as an ``Event ARG'', or eARG: 
an example is given in Fig.~\ref{fig-arg-data-structure}.
Each node in the graph corresponds to an event: therefore the graph
contains both recombination nodes (node \textsf{D}) and common ancestor
nodes (nodes \textsf{E}, \textsf{F}, and \textsf{G}) corresponding to their
respective event types. Edges represent lineages,
and those lineages are modified by the events that they pass through.

@@

\node [nodelabel,anchor=north west] at ($(s3) + (0,0)$) {$x$};
\foreach \u/\lab in {s0/$\textsf{A}$, s1/$\textsf{B}$, s2/$\textsf{C}$} \node[nodelabel,anchor=north] at (\u) {\lab};
\foreach \u/\lab in {s4/$\textsf{E}$} \node[nodelabel,anchor=south west] at (\u) {\lab};
\foreach \u/\lab in {s5/$\textsf{F}$} \node[nodelabel,anchor=south east] at (\u) {\lab};
\foreach \u/\lab in {s3/$\textsf{D}$, s6/$\textsf{G}$} \node[nodelabel,anchor=south] at (\u) {\lab};

@@

\node [nodelabel,anchor=north west] at ($(-10,5)$) {
\begin{tabular}{c|c}
% \multicolumn{2}{c}{Breakpoint}\\
node & breakpoint\\
\hline
$\textsf{A}$ & $\varnothing$ \\
$\textsf{B}$ & $\varnothing$ \\
$\textsf{C}$ & $\varnothing$ \\
$\textsf{D}$ & $x$ \\
$\textsf{E}$ & $\varnothing$ \\
$\textsf{F}$ & $\varnothing$ \\
$\textsf{G}$ & $\varnothing$ \\
\end{tabular}};

\node [nodelabel,anchor=north west] at ($(-4,5)$) {
\begin{tabular}{l}
Edges\\
\hline
$(\textsf{A}, \textsf{E})$ \\
$(\textsf{B}, \textsf{D})$ \\
$(\textsf{C}, \textsf{F})$ \\
$(\textsf{D}, \textsf{E})$ \\
$(\textsf{D}, \textsf{F})$ \\
$(\textsf{E}, \textsf{G})$ \\
$(\textsf{F}, \textsf{G})$ \\
\end{tabular}};

\node [nodelabel,anchor=north west] at ($(7,5)$) {
\begin{tabular}{ll}
Edges\\
\hline
$(\textsf{A}, \textsf{E})$ & $(0, L]$ \\
$(\textsf{B}, \textsf{D})$ & $(0, L]$ \\
$(\textsf{C}, \textsf{F})$ & $(0, L]$ \\
$(\textsf{D}, \textsf{E})$ & $(0, x]$ \\
$(\textsf{D}, \textsf{F})$ & $(x, L]$ \\
$(\textsf{E}, \textsf{G})$ & $(0, L]$ \\
$(\textsf{F}, \textsf{G})$ & $(0, L]$ \\
\end{tabular}};

@@

% What do we mean by an EARG, formally?
% NOTE we're using the natural numbers but are using 0 as a node in the
% examples. Probably want to clear this up a bit.
Formally, we can define an eARG as a tuple $(e, \sigma)$, where $e$
is an ordered list of edges defining a directed acyclic graph and
$\sigma: \mathbb{N} \rightarrow \mathbb{N}$
is a function mapping nodes to recombination breakpoints,
such that $\sigma(u) = x$
if $u$ is a recombination event with breakpoint $x$ and
$\sigma(u) = \varnothing$ otherwise. (This formulation is equivalent to stipulating that
a recombination node is of a different ``type'' to other nodes,
with breakpoints only associated with this specific node type).
Each edge $e_j = (c, p)$ describes an edge between
child node $c$ and parent $p$, where $c, p \in \mathbb{N}$.
For each recombination node $u$ we must
have exactly two edges in $e$ such that $u$ is the child.
Note that this description captures only the
graph topology: if we also wish to know the times of events we need
an additional function $t(u): \mathbb{N} \rightarrow \mathbb{R}$
which defines the time of each node.

@@

A final limitation of the eARG encoding illustrated in
Fig.~\ref{fig-arg-data-structure} is the inability to represent
multiple recombinations along a chromosome, gene conversion,
or transmission of multiple chromosomes.
[Sentence saying ``gene conversion is important, actually, cite, cite,
so we should be hoping to infer it some day''. Presumably people
already are in bacterial contexts?]
To allow us to encode data from these important processes we
must either devise workarounds to the event based encoding
(i.e., model gene conversion as two recombination events) or
generalise the model to more appropriately reflect the realities
of modern datasets.

Fortunately, it is straightforward to generalise the classical Griffiths
eARG encoding to both simplify implementations and to encompass any
possible pattern of genetic inheritance.
The only change that we require is to associate
the details of inheritance intervals with \emph{edges} rather than
as breakpoints associated with recombination events and
edge ordering rules. This is illustrated in Fig~\ref{fig-arg-data-structure}B
where this alternative edge annotation formulation is
shown. By associating the interval $(0, x)$ with edge $(3,4)$
and the interval $(x, L)$ with $(3, 5)$ we make the encoding
unambiguous: we can list the edges in any order, and it
will describe the same set of local genealogical trees.
This approach is also more flexible, because any pattern
of inheritance can be described by these intervals. For
example, if a gene conversion event occurred where a segment
between between positions $x$ and $y$ was inherited by
node $3$ from $5$, we would state this by associating
the intervals $\{(0, x], (y, L]\}$ with $(3,4)$ and
$\{(x, y]\}$ with $(3,5)$.
This change is straightforward, involves no loss of information
and (as we see in later sections)
greatly increases both the expressivity and computational
power of ancestral recombination graph data structures.

@@

\begin{figure}
\begin{center}
    \includegraphics[width=\textwidth]{illustrations/arg-in-pedigree}
\end{center}
\caption{\label{fig-arg-in-pedigree}
% TODO caption needs another pass. One the full section has been written
% it'll be clearer what we need to say in the caption vs the text.
The Ancestral Recombination Graph defined in terms of genomes embedded
in a pedigree. (a) Diploid individuals, visualised in a highly inbred pedigree and
labelled $i_1$ to $i_8$, contain both paternal (left) and maternal (right) genomes
labelled \textsf{A} to \textsf{P}. Black lines show transmission paths connecting
genomes in the current generation (\textsf{A} to \textsf{D}) with their ancestors.
Two recombination breakpoints, at positions 2 and 7 mean that genomes \textsf{A}
and \textsf{C} are independent recombinants of the paternal genomes \textsf{E}
and \textsf{F}: in these cases the edge annotations have been shown, indicating the
region of genome transmitted. Genomes are shaded such that where, backwards in time,
they merge into a common ancestor, the merged region is darker; this shows that all
genomic regions are fully coalesced in genome \textsf{N} in the oldest generation.
(b) The equivalent genome ARG, where there is a single type of node (representing
a genome) along with annotations on all edges; annotations showing partial
transmission are highlighted in bold. Recombination, such as that causing the
breakpoint at position 7, is associated with two parent genomes (\textsf{E} and
\textsf{F}) and a child genome (the recombinant, \textsf{C}).
(c) The edge-annotated genome ARG in (b) fully defines the
local relationships along the genome, as shown in the local
or ``marginal'' trees.
}
\end{figure}

To distinguish from the classical event-based encodings,
we call this encoding of an ARG data
structure a ``Genome ARG'', or gARG.
Fig~\ref{fig-arg-in-pedigree} illustrates the basic elements of a gARG.
Fig.~\ref{fig-arg-in-pedigree}a shows the genomes
embedded in an inbred diploid pedigree, in which individuals occur at
discrete generations and carry two genomes. Each individual (labelled
$i_1$ to $i_8$) therefore has a paternal genome (on the right) and a maternal one
(on the left), although the pedigree is only shown for concreteness, and
there is no inherent limitation on mating system, ploidy, or age structure.
A genome is inherited from one of the individual's parents,
and may be the recombined product of that parent's two genomes.
For example, individual $i_1$ has two genomes \textsf{A} and \textsf{B},
inherited from parents $i_3$ and $i_4$. Genome \textsf{A} is the product of
recombining $i_3$'s two genomes \textsf{E} and \textsf{F} at position 2,
whereas \textsf{B} was inherited directly from $i_4$'s genome \textsf{G} without
recombination. Fig.~\ref{fig-arg-in-pedigree}b shows the bare gARG, illustrating
that the patterns of
transmission between genomes can be followed without necessarily knowing
how they are combined within individuals. This ``genome-perspective''
simply depicts the ancestry of the sample genomes $A$--$D$. The graph
topology along with the genome coordinate annotations on the
edges then uniquely defines the trees at every position
along the genome (Fig.~\ref{fig-arg-in-pedigree}c).

@@

The focus on genomes rather than events has some interesting
and important ramifications. Firstly, a gARG may contain nodes that
do not correspond to any particular event in the ancestry of the samples.
For example, node \textsf{H} is the direct ancestor of \textsf{D} in
Fig~\ref{fig-arg-in-pedigree}, and is not the product of either a
coalescence or a recombination. Such nodes would usually be removed
from the graph so that \textsf{D} descends directly from \textsf{K} (see
the ``\nameref{ARG_simplification}'' section) but there is no necessity for this from
a representational perspective. This ability can be useful in practice and could be
% Do we need to spell this out?
required if, for example, we have a deeply sequenced pedigree of haplodiploids.
% e.g. Honeybees? Be great to cite an example!
% Note by Yan - if we actually have the sequence, then this turns into
% a sample node, so would be normally included anyway. Perhaps, therefore
% we should talk about these "pass though" nodes in the context
% of internal samples.

Secondly, unlike in an eARG, nodes in a gARG are not classified into different
types. Indeed a single node can have both multiple children
and multiple parents. This would be the case in Fig~\ref{fig-arg-in-pedigree} if,
for instance, node \textsf{A} were to have two children.
% It is a shame that this is not actually shown in fig 3, but I don't think
% we can do this unless we have 5 generations
Moreover, a single node can have more than 2 children, and (as we shall see in
the ``\nameref{ARG_simplification}'' section, more than 2 parents. Another way to
state this is that transitions between nodes in a gARG can encapsulate multiple
events in an eARG.

The final point deals with the way in which recombination is represented, and is
closely linked to which genomes are chosen as the nodes in a gARG.
We could potentially choose genomes at any point in the life cycle: for instance,
we could chose to represent gametes as ARG nodes. However, the most intuitive
choice is where genomes represent diploid individuals in each generation, as in
Fig~\ref{fig-arg-in-pedigree}a. In this case recombination
is not represented by a single ``recombination node'' (as in an eARG), but with
two parent genomes between which recombination takes place (e.g.
nodes \textsf{E} and \textsf{F}), and a
single resulting ``recombinant'' genome (e.g. \textsf{C}). The
edges that link this recombinant with its two parents carry the information
concerning breakpoint location that would normally be present on the
``recombination nod'' in an eARG. For example, node \textsf{C} in
Fig~\ref{fig-arg-in-pedigree} inherits genome positions 0 to 7 from \textsf{E}
and positions 7 to 10 from \textsf{F}, indicating that the breakpoint is at position 7.

@@

\begin{figure}
\centering
\includegraphics[width=\textwidth]{illustrations/ancestry-resolution}
\caption{\label{fig-ancestry-resolution}
The \citet[][fig. 1]{wiuf1999recombination} example graph as a gARG, with contrasting
edge annotation schemes.
(a) Edge annotations that reflect immediate inheritance, i.e. transmission of all
genetic material between parent and child. All edges span the full genome from position 0 to 7,
apart from those linking a recombinant and its two parents (bold).
(b) Sample-resolved edge annotations: the Hudson simplification algorithm has been used to
shorten edge spans to reflect only the upwards transmission of genetic
material from the samples \textsf{A}, \textsf{B}, and \textsf{C}. The depiction of the genome within
each node is shaded as in figure \ref{fig-arg-in-pedigree}a to show
coalescing regions of genome back to the grand MRCA genome \textsf{W}.
In this case, using line widths to reflect the genomic span of each edge can
help to visualise the relative importance of different transmission pathways.
}
\end{figure}

@@

\begin{figure}
\centering
\vspace{5em}
\includegraphics[width=\linewidth]{illustrations/simplification}
\caption{\label{fig-simplification}
ARG simplification. In all cases the graph visualization is shown on the left
and the equivalent marginal trees on the right.
(A) A gARG simulated from a diploid Wright-Fisher
model. Note that the genome \textsf{K} has three children, and the genomes
\textsf{E} and \textsf{G} simultaneously have multiple parents and multiple children;
neither could be directly represented in a Griffiths-like eARG encoding
(see section XXX). The penultimate local tree is highlighted to aid discussion (see section YYY).
(B) Simplified to remove all
% trying out this terminology - can define in the text
1-connected graph components (e.g., diamonds such as \textsf{JLMN}).
(C) Remove nodes that are unary in all local trees. This removes both recombinant nodes
that have only one child (e.g. \textsf{N}) and nodes that have more
than one child (i.e. are ``common ancestors'') but never represent coalescences
in the local trees (e.g. \textsf{R}).
(D) Rewrite edges to bypass any unary nodes in local trees. Note that this loses lineage
information such as the fact that \textsf{K} is connected to \textsf{A} via \textsf{E}.
}
\end{figure}

@@

The ideas of ``meaningful'' local tree topologies revolve around the
presence of ``unary'' nodes: a node in the tree that has exactly
one child. Fig.~\ref{fig-simplification} show a series of successive steps of
simplification, starting with a complete gARG simulated under a Wright-Fisher
model, shown both as a graph on the left, and the equivalent set of 6
local trees on the right (Fig.~\ref{fig-simplification}A). Consider the path above node
\textsf{A} in the penultimate tree, highlighted in yellow and covering genomic positions
60 to 70. On this path, before $A$ meets its common ancestor with the other samples at
node \textsf{Q}, it passes through  \textsf{E},  \textsf{J},  \textsf{M},  \textsf{N} and 
\textsf{P}, which in this region of the genome are all unary nodes. This reflects the
passage of this span of genome through the ARG towards the root (see the
``\nameref{Ancestry_resolution}'' section).
Such unary nodes in a local tree are not usually considered important,
because they provide very little information and are not
detectable by information from the tree. % weak
The nodes in a tree usually mark the points where samples find a
 \emph{common} ancestor, such as node \textsf{Q} in the highlighted
tree. For many purposes, we would therefore consider a path directly
joining sample \textsf{A} to node \textsf{Q} to be of equivalent status to the path
containing the intermediate unary nodes. Simplification is the process of removing
these less informative and (for most computational purposes) inefficient nodes
from the local trees and the corresponding annotated graph.

The first type of simplification that we can perform is to remove
graph topology that is invisible to the samples. The most well
known example of such topology is a so-called
``diamond''~\citep{rasmussen2014genome}
in which the two parent nodes of a recombination immediately
join again into a common ancestor (e.g. \textsf{J}, \textsf{L}, \textsf{M}
and  \textsf{N} in Fig.~\ref{fig-simplification}A).
Unless we are specifically
interested in the recombination event or ancestral genomes,
there is no information in this topology and the diamond can be
replaced by a single edge. More generally, any
subgraph which is singly-connected in both the leafward and
rootward direction (a ``super-diamond'') is non-identifiable and can be
replaced by a single edge. This definition includes the case
of a single node that has one inbound and one outbound edge, such as
nodes \textsf{F} and \textsf{H}.
Fig.~\ref{fig-simplification}B shows the result of this type of
graph topology simplification.

>>>>

% NOTE: it might be helpful to have a drawing of the ARG process
% in-progress here to help us talk about it. What is the full state
% that is remembered as we go through the process? It does help to think
% about the process as choosing edges to merge, not nodes.
% We could also refer to Fig.~\ref{fig-arg-data-structure}B if we want a simple picture
% to go with this
In the Griffiths formulation, each edge in the graph corresponds to an extant
lineage and nodes are events in the process (the initial $n$ leaf nodes are
``sampling'' events). Common ancestor events occur at rate $\binom{k}{2}$ when there
are $k$ lineages present. We choose two lineages (edges) uniformly, and merge them
into a common ancestor lineage. Recombination events happen at a rate of
$k \rho$. Here, we choose a lineage (edge) uniformly, and a
breakpoint $0 < x < m$ uniformly on its genome. We terminate the edge at a
node, record the breakpoint and start two new edges from this node. The process
then continues until there is only one lineage left (the Grand Most Recent
Common Ancestor, GMRCA), which is guaranteed to
happen in finite time because of the linear vs quadratic rates of branching
and coalescing. Later, we will show how the graph structure and breakpoints
associated with
recombination nodes provides sufficient information to later recover the marginal
trees and calculate sampling probabilities. However, this process traces out
a larger graph than the Hudson algorithm (see the ``\nameref{Ancestry_resolution}'' section),
and is consequently considerably less efficient \citep{wiuf1999ancestry}.

@@

\section*{The Event ARG data structure}\label{eARG}
% When did we move from the ARG is a process to the ARG is a data structure?
% What was the historical progression from one to the other?

@@

% What do we mean by an EARG, informally?
The Griffiths ARG encoding is an elegant and mathematically
economical description of the coalescent with recombination, whose
fundamental units are the  \emph{events} that correspond directly
to those that occur in the associated stochastic process. We will refer
to this ARG encoding as an ``Event ARG'', or eARG: 
an example is given in Fig.~\ref{fig-arg-data-structure}B.
Each node in the graph corresponds to an event: therefore the graph
contains both recombination nodes (node \textsf{d}) and common ancestor
nodes (nodes \textsf{e}, \textsf{f}, and \textsf{g}) corresponding to their
respective event types. Edges represent lineages,
and those lineages are modified by the events that they pass through.

@@

\node [anchor=north west] at (-1.5,6) {B};
\node [nodelabel,anchor=north west] at ($(s3) + (0,0)$) {$x$};
\foreach \u/\lab in {s0/$\textsf{a}$, s1/$\textsf{b}$, s2/$\textsf{c}$} \node[nodelabel,anchor=north] at (\u) {\lab};
\foreach \u/\lab in {s4/$\textsf{e}$} \node[nodelabel,anchor=south west] at (\u) {\lab};
\foreach \u/\lab in {s5/$\textsf{f}$} \node[nodelabel,anchor=south east] at (\u) {\lab};
\foreach \u/\lab in {s3/$\textsf{d}$, s6/$\textsf{g}$} \node[nodelabel,anchor=south] at (\u) {\lab};

@@

\node [anchor=north west] at (-14,6) {A};
\node [nodelabel,anchor=north west] at ($(-13,5)$) {
\begin{tabular}{c|c}
% \multicolumn{2}{c}{Breakpoint}\\
node & breakpoint\\
\hline
$\textsf{a}$ & $\varnothing$ \\
$\textsf{b}$ & $\varnothing$ \\
$\textsf{c}$ & $\varnothing$ \\
$\textsf{d}$ & $x$ \\
$\textsf{e}$ & $\varnothing$ \\
$\textsf{f}$ & $\varnothing$ \\
$\textsf{g}$ & $\varnothing$ \\
\end{tabular}};

\node [nodelabel,anchor=north west] at ($(-7,5)$) {
\begin{tabular}{l}
Edges\\
\hline
$(\textsf{a}, \textsf{e})$ \\
$(\textsf{b}, \textsf{d})$ \\
$(\textsf{c}, \textsf{f})$ \\
$(\textsf{d}, \textsf{e})$ \\
$(\textsf{d}, \textsf{f})$ \\
$(\textsf{e}, \textsf{g})$ \\
$(\textsf{f}, \textsf{g})$ \\
\end{tabular}};

\node [anchor=north west] at (9,6) {C};
\node [nodelabel,anchor=north west] at ($(10,5)$) {
\begin{tabular}{ll}
Edges\\
\hline
$(\textsf{a}, \textsf{e})$ & $(0, L]$ \\
$(\textsf{b}, \textsf{d})$ & $(0, L]$ \\
$(\textsf{c}, \textsf{f})$ & $(0, L]$ \\
$(\textsf{d}, \textsf{e})$ & $(0, x]$ \\
$(\textsf{d}, \textsf{f})$ & $(x, L]$ \\
$(\textsf{e}, \textsf{g})$ & $(0, L]$ \\
$(\textsf{f}, \textsf{g})$ & $(0, L]$ \\
\end{tabular}};

@@

% What do we mean by an eARG, formally?
Formally, we can define an eARG as a tuple $(e, \sigma)$, where $e$
is an ordered list of edges defining a directed acyclic graph and
$\sigma: \mathbb{N} \rightarrow \mathbb{N}$
is a function mapping nodes to recombination breakpoints,
such that $\sigma(u) = x$
if $u$ is a recombination event with breakpoint $x$ and
$\sigma(u) = \varnothing$ otherwise. (This formulation is equivalent to stipulating that
a recombination node is of a different ``type'' to other nodes,
with breakpoints only associated with this specific node type).
Each edge $e_j = (c, p)$ describes an edge between
child node $c$ and parent $p$, where $c, p \in \mathbb{N}$.
For each recombination node $u$ we must
have exactly two edges in $e$ such that $u$ is the child.
Note that this description captures only the
graph topology: if we also wish to know the times of events we need
an additional function $t(u): \mathbb{N} \rightarrow \mathbb{R}$
which defines the time of each node.

@@

A final limitation of the eARG encoding illustrated in
Fig.~\ref{fig-arg-data-structure}A is the inability to represent
multiple recombinations along a chromosome, gene conversion,
or transmission of multiple chromosomes.
[Sentence saying ``gene conversion is important, actually, cite, cite,
so we should be hoping to infer it some day''. Presumably people
already are in bacterial contexts?]
To allow us to encode data from these important processes we
must either devise workarounds to the event based encoding
(i.e., model gene conversion as two recombination events) or
generalise the model to more appropriately reflect the realities
of modern datasets.

Fortunately, it is straightforward to generalise the classical Griffiths
eARG encoding to both simplify implementations and to encompass any
possible pattern of genetic inheritance.
The only change that we require is to associate
the details of inheritance intervals with \emph{edges} rather than
as breakpoints associated with recombination events and
edge ordering rules. This is illustrated in Fig~\ref{fig-arg-data-structure}C
where this alternative edge annotation formulation is
shown. By associating the interval $(0, x)$ with edge $(3,4)$
and the interval $(x, L)$ with $(3, 5)$ we make the encoding
unambiguous: we can list the edges in any order, and it
will describe the same set of local genealogical trees.
This approach is also more flexible, because any pattern
of inheritance can be described by these intervals. For
example, if a gene conversion event occurred where a segment
between between positions $x$ and $y$ was inherited by
node $3$ from $5$, we would state this by associating
the intervals $\{(0, x], (y, L]\}$ with $(3,4)$ and
$\{(x, y]\}$ with $(3,5)$.
This change is straightforward, involves no loss of information
and (as we see in later sections)
greatly increases both the expressivity and computational
power of ancestral recombination graph data structures.

@@

\begin{figure}
\begin{center}
    \includegraphics[width=\textwidth]{illustrations/arg-in-pedigree}
\end{center}
\caption{\label{fig-arg-in-pedigree}
% TODO caption needs another pass. One the full section has been written
% it'll be clearer what we need to say in the caption vs the text.
The Ancestral Recombination Graph defined in terms of genomes embedded
in a pedigree. (A) Diploid individuals, visualised in a highly inbred pedigree and
labelled $I_1$ to $I_8$, contain both paternal (left) and maternal (right) genomes
labelled \textsf{a} to \textsf{p}. Black lines show transmission paths connecting
genomes in the current generation (\textsf{a} to \textsf{d}) with their ancestors.
Two recombination breakpoints, at positions 2 and 7 mean that genomes \textsf{a}
and \textsf{c} are independent recombinants of the paternal genomes \textsf{e}
and \textsf{f}: in these cases the edge annotations have been shown, indicating the
region of genome transmitted. Genomes are shaded such that where, backwards in time,
they merge into a common ancestor, the merged region is darker; this shows that all
genomic regions are fully coalesced in genome \textsf{n} in the oldest generation.
(B) The equivalent genome ARG, where there is a single type of node (representing
a genome) along with annotations on all edges; annotations showing partial
transmission are highlighted in bold. Recombination, such as that causing the
breakpoint at position 7, is associated with two parent genomes (\textsf{e} and
\textsf{f}) and a child genome (the recombinant, \textsf{c}).
(C) The edge-annotated genome ARG in (B) fully defines the
local relationships along the genome, as shown in the local
or ``marginal'' trees.
}
\end{figure}

To distinguish from the classical event-based encodings,
we call this encoding of an ARG data
structure a ``Genome ARG'', or gARG.
Fig~\ref{fig-arg-in-pedigree} illustrates the basic elements of a gARG.
Fig.~\ref{fig-arg-in-pedigree}A shows the genomes
embedded in an inbred diploid pedigree, in which individuals occur at
discrete generations and carry two genomes. Each individual (labelled
$I_1$ to $I_8$) therefore has a paternal genome (on the right) and a maternal one
(on the left), although the pedigree is only shown for concreteness, and
there is no inherent limitation on mating system, ploidy, or age structure.
A genome is inherited from one of the individual's parents,
and may be the recombined product of that parent's two genomes.
For example, individual $I_1$ has two genomes \textsf{a} and \textsf{b},
inherited from parents $I_3$ and $I_4$. Genome \textsf{a} is the product of
recombining $I_3$'s two genomes \textsf{e} and \textsf{e} at position 2,
whereas \textsf{b} was inherited directly from $I_4$'s genome \textsf{g} without
recombination. Fig.~\ref{fig-arg-in-pedigree}B shows the bare gARG, illustrating
that the patterns of
transmission between genomes can be followed without necessarily knowing
how they are combined within individuals. This ``genome-perspective''
simply depicts the ancestry of the sample genomes \textsf{a}--\textsf{d}. The graph
topology along with the genome coordinate annotations on the
edges then uniquely defines the trees at every position
along the genome (Fig.~\ref{fig-arg-in-pedigree}C).

@@

The focus on genomes rather than events has some interesting
and important ramifications. Firstly, a gARG may contain nodes that
do not correspond to any particular event in the ancestry of the samples.
For example, node \textsf{h} is the direct ancestor of \textsf{d} in
Fig~\ref{fig-arg-in-pedigree}, and is not the product of either a
coalescence or a recombination. Such nodes would usually be removed
from the graph so that \textsf{d} descends directly from \textsf{k} (see
the ``\nameref{ARG_simplification}'' section) but there is no necessity for this from
a representational perspective. This ability can be useful in practice and could be
% Do we need to spell this out?
required if, for example, we have a deeply sequenced pedigree of haplodiploids.
% e.g. Honeybees? Be great to cite an example!
% Note by Yan - if we actually have the sequence, then this turns into
% a sample node, so would be normally included anyway. Perhaps, therefore
% we should talk about these "pass though" nodes in the context
% of internal samples.

Secondly, unlike in an eARG, nodes in a gARG are not classified into different
types. Indeed a single node can have both multiple children
and multiple parents. This would be the case in Fig~\ref{fig-arg-in-pedigree} if,
for instance, node \textsf{a} were to have two children.
% It is a shame that this is not actually shown in fig 3, but I don't think
% we can do this unless we have 5 generations
Moreover, a single node can have more than 2 children, and (as we shall see in
the ``\nameref{ARG_simplification}'' section), more than 2 parents. Another way to
state this is that transitions between nodes in a gARG can encapsulate multiple
events in an eARG.

The final point deals with the way in which recombination is represented, and is
closely linked to which genomes are chosen as the nodes in a gARG.
We could potentially choose genomes at any point in the life cycle: for instance,
we could chose to represent gametes as ARG nodes. However, the most intuitive
choice is where genomes represent diploid individuals in each generation, as in
Fig~\ref{fig-arg-in-pedigree}A. In this case recombination
is not represented by a single ``recombination node'' (as in an eARG), but with
two parent genomes between which recombination takes place (e.g.
nodes \textsf{e} and \textsf{f}), and a
single resulting ``recombinant'' genome (e.g. \textsf{c}). The
edges that link this recombinant with its two parents carry the information
concerning breakpoint location that would normally be present on the
recombination node in an eARG. For example, node \textsf{c} in
Fig~\ref{fig-arg-in-pedigree} inherits genome positions 0 to 7 from \textsf{e}
and positions 7 to 10 from \textsf{f}, indicating that the breakpoint is at position 7.

@@

\begin{figure}
\centering
\includegraphics[width=\textwidth]{illustrations/ancestry-resolution}
\caption{\label{fig-ancestry-resolution}
The \citet[][fig. 1]{wiuf1999recombination} example graph as a gARG, with contrasting
edge annotation schemes.
(A) Edge annotations that reflect immediate inheritance, i.e. transmission of all
genetic material between parent and child. All edges span the full genome from position 0 to 7,
apart from those linking a recombinant and its two parents (bold).
(B) Sample-resolved edge annotations: the Hudson simplification algorithm has been used to
shorten edge spans to reflect only the upwards transmission of genetic
material from the samples \textsf{a}, \textsf{b}, and \textsf{c}. The depiction of the genome within
each node is shaded as in figure \ref{fig-arg-in-pedigree}a to show
coalescing regions of genome back to the grand MRCA genome \textsf{w}.
In this case, using line widths to reflect the genomic span of each edge can
help to visualise the relative importance of different transmission pathways.
}
\end{figure}

@@

\begin{figure}
\centering
\vspace{5em}
\includegraphics[width=\linewidth]{illustrations/simplification}
\caption{\label{fig-simplification}
ARG simplification. In all cases the graph visualization is shown on the left
and the equivalent marginal trees on the right.
(A) A gARG simulated from a diploid Wright-Fisher
model. Note that the genome \textsf{k} has three children, and the genomes
\textsf{e} and \textsf{g} simultaneously have multiple parents and multiple children;
neither could be directly represented in a Griffiths-like eARG encoding
(see the ``\nameref{eARG}'' section). The penultimate local tree is highlighted to aid discussion.
(B) Simplified to remove all
% trying out this terminology - can define in the text
1-connected graph components (e.g., diamonds such as \textsf{jlnm}).
(C) Remove nodes that are unary in all local trees. This removes both recombinant nodes
that have only one child (e.g. \textsf{n}) and nodes that have more
than one child (i.e. are ``common ancestors'') but never represent coalescences
in the local trees (e.g. \textsf{r}).
(D) Rewrite edges to bypass any unary nodes in local trees. Note that this loses lineage
information such as the fact that \textsf{k} is connected to \textsf{a} via \textsf{e}.
}
\end{figure}

@@

The ideas of ``meaningful'' local tree topologies revolve around the
presence of ``unary'' nodes: a node in the tree that has exactly
one child. Fig.~\ref{fig-simplification} show a series of successive steps of
simplification, starting with a complete gARG simulated under a Wright-Fisher
model, shown both as a graph on the left, and the equivalent set of 6
local trees on the right (Fig.~\ref{fig-simplification}A). Consider the path above node
\textsf{a} in the penultimate tree, highlighted in yellow and covering genomic positions
60 to 70. On this path, before $A$ meets its common ancestor with the other samples at
node \textsf{q}, it passes through  \textsf{e},  \textsf{j},  \textsf{m},  \textsf{n} and 
\textsf{p}, which in this region of the genome are all unary nodes. This reflects the
passage of this span of genome through the ARG towards the root (see the
``\nameref{Ancestry_resolution}'' section).
Such unary nodes in a local tree are not usually considered important,
because they provide very little information and are not
detectable by information from the tree. % weak
The nodes in a tree usually mark the points where samples find a
 \emph{common} ancestor, such as node \textsf{q} in the highlighted
tree. For many purposes, we would therefore consider a path directly
joining sample \textsf{a} to node \textsf{q} to be of equivalent status to the path
containing the intermediate unary nodes. Simplification is the process of removing
these less informative and (for most computational purposes) inefficient nodes
from the local trees and the corresponding annotated graph.

The first type of simplification that we can perform is to remove
graph topology that is invisible to the samples. The most well
known example of such topology is a so-called
``diamond''~\citep{rasmussen2014genome}
in which the two parent nodes of a recombination immediately
join again into a common ancestor (e.g. \textsf{j}, \textsf{l}, \textsf{m}
and  \textsf{n} in Fig.~\ref{fig-simplification}A).
Unless we are specifically
interested in the recombination event or ancestral genomes,
there is no information in this topology and the diamond can be
replaced by a single edge. More generally, any
subgraph which is singly-connected in both the leafward and
rootward direction (a ``super-diamond'') is non-identifiable and can be
replaced by a single edge. This definition includes the case
of a single node that has one inbound and one outbound edge, such as
nodes \textsf{f} and \textsf{h}.
Fig.~\ref{fig-simplification}B shows the result of this type of
graph topology simplification.

Simplifying away diamonds will remove many unary nodes from the
local trees, but there can still be nodes that are unary in all
of the local trees. In particular, a node can represent a recombinant
with multiple parents in the graph but only a single child (e.g. node \textsf{n}
in Fig.~\ref{fig-simplification}B), or can represent a common ancestor with
multiple children in the graph but in which no local coalescence takes place
(node \textsf{r} in Fig.~\ref{fig-simplification}B).
% point out that this is why Hudson refers to CA nodes, not coalescence nodes
Such nodes are not singly connected in the graph, but are nevertheless unary in
all of the local trees in Fig.~\ref{fig-simplification}B. The operation to remove them,
the results of which are shown in Fig.~\ref{fig-simplification}C,
therefore requires knowledge not just of the graph topology, but also of the
edge annotations. Removal of recombinant nodes can produce graph nodes with
more than 2 parents (e.g. node \textsf{e}); likewise, removal of
common ancestor but non-coalescent nodes can produce graph nodes with
more than 2 children (e.g. node \textsf{s}). These represent a ``stacking up'' of multiple
often indistinguishable events into a single node, something which is impossible
to encode in an eARG.

The remaining nodes are MRCAs of some subset of the samples
at \emph{some} positions along the genome. We still have
some unary nodes in the local trees, but these nodes will
correspond to a coalescence in at least one other
local tree. For example, node  \textsf{k} is unary in the second tree
of Fig.~\ref{fig-simplification}C, but is either binary
or ternary in all subsequent trees (recall this is a Wright-Fisher
simulation). The final level of simplification is to alter the edge annotations
such that, although no nodes are removed from the graph, \emph{all}
unary nodes disappear from the local trees (Fig.~\ref{fig-simplification}D).
It is common for this operation to remove nodes above the top MRCA node
in the tree, resulting in many more root nodes (here, \textsf{q} and \textsf{k}
become roots). [And continue]
% Note that although this last stage produces much simpler local trees, by
% removing information about the exact paths taken by lineages through
% the graph, it loses potentially important information about shared edges
% between trees. It is not clear whether (c) or (d) leads to a more compact
% representation of the ARG in terms of number of edges.
